---
title: 'Reproducible report: <br/> Meta-analysis of relationships in soybean white mold epidemic data'
date: "`r Sys.Date()`"
output:
  html_document:
    css: my-style.css
    depth: 5
    number_sections: no
    theme: flatly
    toc: yes
    toc_float:
      collapsed: yes
      smooth_scroll: yes
  pdf_document:
    toc: yes
---

<link href="https://fonts.googleapis.com/css?family=Oswald" rel="stylesheet">




```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(width=200)
```


# The article

This report describes all steps for reproducing the analysis of data from multiple field trial, using meta-analytic approaches, conducted in an article published in Plant Pathology:

```{r, message=FALSE, warning=FALSE, include=FALSE}
library(rcrossref)
citation <- cr_cn(dois = "10.1111/ppa.12590", format = "text", style = "apa")
```

> `r citation`

<<<<<<< HEAD
In the [article](Lehner-etal-white-mold-PP.pdf), two relationships were studied: soybean white mold incidence (inc, %) and soybean yield (yld, kg/ha) and incidence and sclerotia weight (scl, kg). The data were collected at 35 studies was available in tables, one for each study, published in a scientific summary report. [Download](http://ainfo.cnptia.embrapa.br/digital/bitstream/item/101371/1/Ensaios-cooperativos-de-controle-quimico-de-mofo-branco-na-cultura-da-soja-safras-2009-a-2012.pdf) the report (PT language). The two relationships were summarized using [meta-analytic](http://apsjournals.apsnet.org/doi/abs/10.1094/PHYTO-03-10-0069) models using three effect-sizes: 1) Fisher's Z (from transforming Pearson's r); 2) intercept and 3) slopes of a random-coefficients models fitted to the data. We followed the procedures  used in previous studies in plant pathology. However, for summarizing the slopes and intercepts, we used a mixed, also know as random-coeficients, model as described in Madden and Paul (2009) who used SAS. Here, we conducted the analysis in `R` with the `lmer` package as described in this [tutorial](http://www.metafor-project.org/doku.php/tips:two_stage_analysis). 
=======
In the [article](Lehner-etal-white-mold-PP.pdf), two relationships were studied: soybean white mold incidence (inc, %) and soybean yield (yld, kg/ha) and incidence and sclerotia weight (scl, g/ha). The data were obtained from a scientific [report](http://ainfo.cnptia.embrapa.br/digital/bitstream/item/101371/1/Ensaios-cooperativos-de-controle-quimico-de-mofo-branco-na-cultura-da-soja-safras-2009-a-2012.pdf) (PT language) on fungicide efficacy evaluated in 35 trials conducted across several locations and 4-year period in Brazil.  The data was organized in tables reporting the mean values of the three variables of interest, one table for each trial.
>>>>>>> gh-pages

The two relationships were summarized using [meta-analytic](http://apsjournals.apsnet.org/doi/abs/10.1094/PHYTO-03-10-0069) models using three effect-sizes: 1) Fisher's Z (from transforming Pearson's r); 2) intercept and 3) slopes of a random-coefficients models fitted to the data. We followed the procedures detailed in previous studies in plant pathology. For summarizing the correlation coefficients, we calculate Fisher's Z and fitted random-effects and mixed-effects model, the latter for testing the effect of moderator variables. For summarizing the slopes and intercepts, we used a mixed, or a random-coeficients, model as described in Madden and Paul (2009) who used SAS. Here, we demonstrate the analysis in `R` using the `lmer` package as described in this [tutorial](http://www.metafor-project.org/doku.php/tips:two_stage_analysis). Similar to the tutorial, the estimates were similar between the two-stage and the mixed model approach, and so we decided to use the latter approach.

# This report

My goal is to demonstrate, using R programming, each step of the analyses, from data preparation to presentation of results, which can be reproduced either by myself or other people interested in the same topic. Instead of giving away all the data and my (messy!) original code, I thougt that investing time to prepare this report can contribute to popularize the use of meta-analytic approaches in plant pathology. The data, codes, pre-print version of the article and supplementary materials were also made available at this GitHub [repository](https://github.com/emdelponte/paper-white-mold-meta-analysis). 

The report is the html output of an [R Markdown](http://rmarkdown.rstudio.com) file prepared with the RStudio IDE for R. As much as possible I used the pipe operator `%>%`  of the [`magrittr` package](https://cran.r-project.org/web/packages/magrittr/vignettes/magrittr.html), which really makes the code easier to write and understand. The plots were prepared using both the base R graphics and ggplot2, whichever was more convenient. Most of them are simple versions for a quick visualization, not actually formatted for final publishing.


# Data import 
First, we need to load the packages.

```{r, message=FALSE, warning=FALSE}
library(tidyverse); library(broom); library(tidyr);
library(cowplot); library(tibble); library(knitr); library(nlme)

```

The raw data was organized in the long or [tidy format](http://garrettgman.github.io/tidying/) where each treatment (observation) in a fungicide trial (hereafter study) is placed in its row and each variable in its column. Let's no load the data and group by study for dataframes separated for each relationship.

```{r, message=FALSE, warning=FALSE}
# inc-yld
dat_yld <- read_csv("dat-white-mold-br.csv") %>%   
 group_by(study)
# inc-scl
dat_scl <- read_csv("dat-white-mold-br.csv") %>%
na.omit(dat_scl)%>% # some trials did not have sclerotia data, and we omit them here
 group_by(study)
```

See the structure of the dataset and all variables types (scroll the content to the left). The full dataset in `csv` format can be downloaded here: [dat-white-mold-br.csv](dat-white-mold-br.csv) 

```{r}
dat_yld
```


# Data visualization

## Raw data

The histograms below summarize the distribution of the three variables used to calculate the three effect-sizes for the meta-analysis. We can see that the two disease-related variables (inc and scl) are skewed to the left and yield data are more normally distributed.

```{r, fig.height=3, fig.width=12, message=FALSE, warning=FALSE}
par(mfrow= c(1,3))
hist(dat_yld$inc, main = "Incidence")
hist(dat_yld$yld, main =  "Yield")
hist(dat_yld$scl, main = "Sclerotia weight")

```

## Individual regressions

Let's now visualize the two relationships of interest, conditioned to study, and add a regression line (extended to the full range) for each study with variable number of pair of observation. Note that soybean yield decreases with the increase of white mold incidence. On the other hand, sclerotia weight increase with the increase of disease incidence. The variability in the incidence in a single study was due to differences in fungicide efficacy. Note that 35 and 29 studies were available for the inc-yld and inc-scl, respectively (6 plots are empty for inc-scl because data was not available).

```{r, fig.height=7, fig.width=10, message=FALSE, warning=FALSE}
library(ggplot2)
library(ggthemes)
ggplot(dat_yld, aes(inc, yld))+
       geom_point(shape = 1)+
       stat_smooth(method = lm, fullrange=TRUE, se = F, col = "black")+
       ylab("Yield (kg/ha)")+
       xlab("White mold incidence (%)")+
       theme_minimal()+
       facet_wrap(~ study, ncol = 7, scales = "fixed") 
      

ggplot(dat_yld, aes(inc, scl))+
       geom_point(shape = 1)+
       stat_smooth(method = lm, fullrange=TRUE, se = F, col = "black")+
       ylab("Sclerotia weight(g)")+
       xlab("White mold incidence (%)")+
       theme_minimal()+
       facet_wrap(~ study, ncol = 7, scales = "fixed") 

```



#### Coefficients


```{r, fig.height=3}
 dat_yld %>%
  do(tidy(lm(.$yld ~ .$inc), conf.int=TRUE)) %>% 
 ggplot(aes(estimate))+
 geom_histogram(bins = 10, fill = "grey90",  color="grey50")+
 theme_minimal()+
 facet_wrap(~term, scales = "free_x")
  
```

#### Fitted lines

Let's see all the regression lines in the same plot.


```{r}
ggplot(dat_yld, aes(inc, yld))+
      geom_smooth(method="lm", fullrange=TRUE, se=F, size=0.7, color="grey80", aes(group = factor(study)))+
      geom_point(alpha = 0.5, shape = 1)+
      ylab("Yield (kg/ha)")+
      xlab("White mold incidence (%)")+
      theme_minimal()
```

### Coefficients


```{r, message=FALSE, warning=FALSE}
dat_scl %>% 
 do(tidy(lm(.$scl ~ .$inc), conf.int=TRUE)) %>%
 ggplot(aes(estimate))+
 geom_histogram(bins = 10, fill = "grey90",  color="grey50")+
 theme_minimal()+
 facet_wrap(~term, scales = "free_x")

```

#### Fitted lines

Regression plot.

```{r, message=FALSE, warning=FALSE}
ggplot(dat_scl, aes(inc, scl))+
      geom_smooth(method="lm", fullrange=TRUE, se=F, size=0.7, color="grey80", aes(group = factor(study)))+
      geom_point(alpha = 0.5, shape = 1)+
      ylab("Sclerotia weight (g/ha)")+
      xlab("White mold incidence (%)")+
      ylim(0,10000)+
      theme_minimal()
```



# Meta-analytic models 

The procedures described below are exactly the same for the two studied relationships. For each one, we firstly summarize the Fisher's Z as the effect size for the study of the strenght of the association between white mold incidence and soybean yield. Secondly, we will fit a regression model to individual studies. Finally, we will fit random-coefficients model to estimate the coefficients.

## Correlation coefficient

### Incidence-yield 


Here I use the `dplyr::do` and `tidyr::broom` functions to extract the correlation statistics from each of the studies and assign them to the `cor_yld_inc` dataframe.


```{r, message=FALSE, warning=FALSE}
cor_yld_inc <- dat_yld %>% 
  do(tidy(cor.test(.$inc, .$yld)))
```

Let's extract the first row of each study from the `dat_yld` dataframe and then combine with the new `cor_yld_inc` dataframe that contains the correlation statistics. We will add a new column (n) for the number of data points per study using the `mutate` function.

```{r}
dat_yld2 <- filter(dat_yld, row_number() == 1)
dat_yld3 <- full_join(cor_yld_inc, dat_yld2, by = "study") %>% 
           mutate(n = parameter + 2)  

```

The Fisher's Z was used as effect-size because of its better statistical property than the Pearson's r. We obtain the Fisher's Z and sampling variance of each study with the `escalc` function of the `metafor` package that calculates and adds them to the dataframe. Note that the effec-size and sampling variance are indicated by `yi` and `vi`, the standard notations used in `metafor` when using the `escalc` function. Let's see how the dataframe looks like.
 

```{r, message=FALSE, warning=FALSE}
library(metafor)
dat_yld3 <- escalc(measure = "ZCOR", ri = estimate, ni = n, data = dat_yld3)

head(dat_yld3)
```

#### Overall Fisher's Z

A [random-effects](http://www.metafor-project.org/doku.php/tips:rma.uni_vs_rma.mv) meta-analytiic model was fitted to these data using a maximum likelihood estimator for the amount of heterogeneity. 

```{r, message=FALSE, warning=FALSE}
ma_cor_yld <- rma.uni(yi, vi, method = "ML", data = dat_yld3)
summary(ma_cor_yld)
```

Back-transform z to obtain overall mean r.

```{r, message=FALSE, warning=FALSE}
pred_r <- predict(ma_cor_yld, transf = transf.ztor)
pred_r
```


#### Moderators effect

The random-effects model fitted previously assumes that the heterogeneity in the true correlation coefficients (Fisher's Z) is purely random. However, there may be differences among the individual effects that are (at least in part) related to study-specific variables. These variable can be treated as "moderators" in the model. We considered here: year, region, elevation, or disease and yield levels in the untreated check, as indicative of the disease pressure.

The mixed-effect models test one moderator variable, each at a time, as a fixed effects. The goal was is examine to what extent the moderators included in the model influence the size of the average true effect. The heterogeneity among the true effect-sizes is evaluated based on significance of the Cochran Q test and the *I*<sup>2</sup> index that measures the extent of heterogeneity of the true effect-sizes.

```{r}
# season
ma_cor_yld_year <- rma(yi, vi, sei="",  mods = ~season, method = "ML",   data = dat_yld3)

## region
ma_cor_yld_region <- rma(yi, vi, mods = ~region, method = "ML",data = dat_yld3)

## elevation as continuous
ma_cor_yld_elevation <- rma(yi, vi, mods = ~elevation, method = "ML", data = dat_yld3)

## elevation as category
ma_cor_yld_elevation2 <- rma(yi, vi, mods = ~elevation_class, method = "ML", data = dat_yld3)

## incidence in the check as continuous
ma_cor_yld_inc <- rma(yi, vi, mods = ~inc_check, method = "ML", data = dat_yld3)

## incidence in the check as categorical
ma_cor_yld_inc2 <- rma(yi, vi, mods = ~inc_class, method = "ML", data = dat_yld3)

## yield in the check as continuous
ma_cor_yld_yld <- rma(yi, vi, mods = ~yld_check, method = "ML", data = dat_yld3)

## yield in the check as categorical
ma_cor_yld_yld2 <- rma(yi, vi, mods = ~yld_class, method = "ML", data = dat_yld3)


```

Let's check which variables significantly affected the heterogeneity and how much the variance was accounted for when including the moderator variable. It seems we have only one variable with a marginally significant P-value.

```{r}

table_yld <- frame_data(
~"Moderator", ~"Test of moderator", ~"R2",
"Season", ma_cor_yld_year$QMp,  ma_cor_yld_year$R2,
"Region",  ma_cor_yld_region$QMp, ma_cor_yld_region$R2,
"Elevation categorical", ma_cor_yld_elevation$QMp,  ma_cor_yld_elevation$R2,
"Elevation continuous", ma_cor_yld_elevation2$QMp, ma_cor_yld_elevation2$R2,
"Incidence categorical", ma_cor_yld_inc$QMp, ma_cor_yld_inc$R,
"Yield continous",  ma_cor_yld_yld$QMp, ma_cor_yld_yld$R2,
"Yield categorical", ma_cor_yld_yld2$QMp, ma_cor_yld_yld2$R2
)

kable(table_yld, format = "pandoc", caption = "P-value for the test for the effect of moderator  and the amount of variance (heterogeneity) accounted for by the moderator (R2 statistics)")
```



#### Forest plot 

Let's visualize the Pearson's r by study and the mean r (solid line) and confidence intervals (dashed lines) from back-transforming Z estimated by the random-effects model. This plots was not include in the paper, but is interesting to check the heterogeneity of the correlation coefficients and the estimated mean r from back-transforming Z (solid line).

```{r}
wi    <- 1/sqrt(dat_yld3$vi)
size  <- 0.5 + 3.0 * (wi - min(wi))/(max(wi) - min(wi))
library(ggplot2)
dat_yld3 %>% 
  ggplot(aes(x = study, y = estimate)) + 
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), color="grey50") + 
  geom_point(aes(size = size), shape = 15, color="grey70") +
 geom_hline(yintercept = pred_r$pred, size=0.75)+
 geom_hline(yintercept = c(pred_r$cr.lb, pred_r$cr.ub), linetype="dashed")+ 
  coord_flip()+
  scale_y_reverse(limits = c(0.65,-1.2))+
  labs(x = "Slope") + 
  theme_minimal() + 
  labs(size = "study weight", 
       title = "White mold vs. soybean yield", 
        y = "Estimated r", x = "Study number (reordered)")
```

Follow the same procedures as described above.

```{r, message=FALSE, warning=FALSE}
cor_scl_inc <- dat_yld %>% 
 na.omit(dat_yld)%>%
  do(tidy(cor.test(.$inc, .$scl)))
```

Join the two data frames 

```{r}
dat_scl2 <- filter(dat_yld, row_number() == 1) %>% 
            na.omit(dat_scl)
dat_scl3 <- full_join(cor_scl_inc, dat_scl2, 
                      by = "study") %>% 
           mutate(n = parameter + 2)  
# create variable with the number of data points in the correlation
```

Obtain Fisher's Z.


```{r, message=FALSE, warning=FALSE}
library(metafor)
dat_scl3 <- escalc(measure = "ZCOR", ri = estimate, ni = n, data = dat_scl3)
```

### Incidence-sclerotia

#### Overall Fisher's Z

Fit the random-coefficients model.

```{r, message=FALSE, warning=FALSE}
ma_cor_scl <- rma.uni(yi, vi, method = "ML", data = dat_scl3)
summary(ma_cor_scl)

```

Back-transform Z to r.

```{r, message=FALSE, warning=FALSE}
pred_r_scl <- predict(ma_cor_scl, transf = transf.ztor)
pred_r_scl
```


#### Moderators effect

Same as previously described.

```{r}
# season
ma_cor_scl_year <- rma(yi, vi, mods = ~season, method = "ML",   data = dat_scl3)

## region
ma_cor_scl_region <- rma(yi, vi, mods = ~region, method = "ML",data = dat_scl3)

## elevation as continuous
ma_cor_scl_elevation <- rma(yi, vi, mods = ~elevation, method = "ML", data = dat_scl3)

## elevation as category
ma_cor_scl_elevation2 <- rma(yi, vi, mods = ~elevation_class, method = "ML", data = dat_scl3)

## incidence in the check as continuous
ma_cor_scl_inc <- rma(yi, vi, mods = ~inc_check, method = "ML", data = dat_scl3)

## incidence in the check as categorical
ma_cor_scl_inc2 <- rma(yi, vi, mods = ~inc_class, method = "ML", data = dat_scl3)

## yield in the check as continuous
ma_cor_scl_yld <- rma(yi, vi, mods = ~yld_check, method = "ML", data = dat_scl3)

## yield in the check as categorical
ma_cor_scl_yld2 <- rma(yi, vi, mods = ~yld_class, method = "ML", data = dat_scl3)


```



```{r}

table_scl <- frame_data(
~"Moderator", ~"Test of moderator", ~"R2",
"Season", ma_cor_scl_year$QMp,  ma_cor_scl_year$R2,
"Region",  ma_cor_scl_region$QMp, ma_cor_scl_region$R2,
"Elevation categorical", ma_cor_scl_elevation$QMp,  ma_cor_scl_elevation$R2,
"Elevation continuous", ma_cor_scl_elevation2$QMp, ma_cor_scl_elevation2$R2,
"Incidence categorical", ma_cor_scl_inc$QMp, ma_cor_scl_inc$R,
"Yield continous",  ma_cor_scl_yld$QMp, ma_cor_scl_yld$R2,
"Yield categorical", ma_cor_scl_yld2$QMp, ma_cor_scl_yld2$R2
)

kable(table_yld, format = "markdown", caption = "P-value of the test for the effect of the moderator and the amount of variance (heterogeneity) accounted for by the moderator (R2 statistics)")
```


#### Forest plot

Visualize the correlation coefficients by study and the mean r (solid line) and confidence intervals (dashed lines) from back-transforming Z estimated by the random-coefficients model. The size of the square is inversely proportion to the study's weight in the analysis. 


```{r}
wi    <- 1/sqrt(dat_scl3$vi)
size  <- 0.5 + 3.0 * (wi - min(wi))/(max(wi) - min(wi))
dat_scl3 %>% 
  ggplot(aes(x = study, y = estimate)) + 
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), color="grey50") + 
  geom_point(aes(size = size), shape = 15, color="grey70") +
  geom_hline(yintercept = pred_r_scl$pred, size=0.75)+
  geom_hline(yintercept = c(pred_r_scl$ci.lb, pred_r_scl$ci.ub), linetype="dashed")+ 
  coord_flip()+
  labs(x = "Slope")+ 
  theme_minimal()+ 
  labs(size = "study weight", 
       title = "White mold incidence vs. sclerotia weight", 
        y = "Estimated r", x = "Study number (reordered)")
```


## Mixed effects model

### Incidence yield

We will use the `lmer` function of the `lme4` to fit three different mixed models: random intercepts and slopes, random intercepts only and random slopes. We followed the procedures described in this [tutorial](http://www.metafor-project.org/doku.php/tips:two_stage_analysis) that compares the mixed model with the two-stage modeling approach.

#### Compare models

```{r, message=FALSE, warning=FALSE}
library(lme4)
# random intercept and slopes
mix_yld1  <- lmer(yld ~ inc + (inc |study), data=dat_yld, REML=F)

# random slopes
mix_yld2 <- lmer(yld ~ inc + (1 | inc), data=dat_yld, REML=F)

# random intercepts
mix_yld3 <- lmer(yld ~ inc + (1 |study), data=dat_yld, REML=F)

```

The AIC of the random intercepts and slopes model is the lowest.

```{r, message=FALSE, warning=FALSE}
AIC(mix_yld1, mix_yld2, mix_yld3)
```

#### Summary

```{r}
summary (mix_yld1)
```


#### Goodness of fit
Let's get the p-value for the significance of the slope and also the pseudo R2 (conditional) of the model.

```{r, message=FALSE, warning=FALSE}
# Get the p-values of the significance of the slope
library(car)
Anova(mix_yld1)

library(piecewiseSEM)
sem.model.fits(mix_yld1)
```


```{r}
# plot the pearson's model residuals
plot(mix_yld1,type=c("p","smooth"))

# plot the qq-plot and qq-line
library(lattice)
qqmath(mix_yld1)

# plot of the fitted and observations
plot(fitted(mix_yld1)~dat_yld$yld)
```

#### Variance-covariance

We then proceed with the full model and calculate the variance and covariance.

```{r}
vc <- VarCorr(mix_yld1)
## variance only
as.data.frame(vc, order="lower.tri")

```

#### Predictions

We now extract the BLUEs and calculate the interdecile range.

```{r}
# extract the blups
cc2 <- coef(mix_yld1)$study
cc2 %>% 
 gather("coef", "value") %>% 
 ggplot(aes(x=value))+
 theme_minimal()+
       geom_histogram(bins = 10, fill = "grey80", color = "grey50")+
       facet_wrap(~coef, scales = "free_x")


```


Calculate the interdecile range for the BLUEs of the slopes and intercepts

```{r, message=FALSE, warning=FALSE}

# Intercept
dec90 <- quantile(cc2$`(Intercept)`,probs=c(.9))
dec10 <- quantile(cc2$`(Intercept)`,probs=c(.1))
dec90-dec10

# Slopes
dec90 <- quantile(cc2$inc,probs=c(.9))
dec10 <- quantile(cc2$inc,probs=c(.1))
dec90-dec10
```

#### Effect of moderators

Forthcoming!


### Incidence-sclerotia

We repeat the same procedure as described above.

#### Compare models

```{r, message=FALSE, warning=FALSE}
library(lme4)
# random intercept and slopes
mix_scl1  <- lmer(scl ~ inc + (inc |study), data=dat_scl, REML=F)
mix_scl2 <- lmer(scl ~ inc + (1 | inc), data=dat_scl, REML=F)
mix_scl3 <- lmer(scl ~ inc + (1 |study), data=dat_scl, REML=F)

```



The AIC  shows the random intercepts and slopes model with the lowest value.

```{r, message=FALSE, warning=FALSE}
AIC(mix_scl1, mix_scl2, mix_scl3)
```


#### Summary

```{r}
summary (mix_scl1)
```

#### Goodness of fit

Let's get the p-value for the significance of the slope and also the pseudo R2 (conditional) of the model.

```{r, message=FALSE, warning=FALSE}
# Get the p-values of the significance of the slope
library(car)
Anova(mix_scl1)

library(piecewiseSEM)
sem.model.fits(mix_scl1)
```

Check the goodness of fit of the model.

```{r}
plot(mix_scl1,type=c("p","smooth"))
library(lattice)
qqmath(mix_scl1)
plot(fitted(mix_scl1)~dat_scl$scl)
```

#### Variance-covariance
We then proceed with the full model and calculate the variance and covariance.

```{r}
vc_scl <- VarCorr(mix_scl1)
## variance only
as.data.frame(vc_scl, order="lower.tri")

```

#### Predictions

Obtaint the BLUEs and calculate the interdecile range.

```{r}
# extract the blups
cc2_scl <- coef(mix_scl1)$study
cc2_scl %>% 
 gather("coef", "value") %>% 
 ggplot(aes(x=value))+
 theme_minimal()+
       geom_histogram(bins = 10, fill = "grey80", color = "grey50")+
       facet_wrap(~coef, scales = "free_x")


```


#### Effect of moderators

Forthcoming!






# References

`r cr_cn(dois = "10.1094/PHYTO-06-14-0157-R", format = "text", style = "apa")`

`r cr_cn(dois = "10.1094/PHYTO-99-7-0850", format = "text", style = "apa")`

